---
title: "Assignment 1: Does Marketing Spend Drive Purchase Amounts?"
author: "Inayat Gill (36880817)"
format:
  html: default
  pdf: default
execute: 
  echo: false
  message: false
  warning: false  
---

```{r setup}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(knitr)
library(scales) 

```

## Introduction 

Every quarter, marketing teams face the same critical question: is our spending actually working? This report looks at a personalised campaign by an online retailer, where each customer received a different level of spend through ads, emails, and coupons. The goal is to understand whether spending more on a customer leads to bigger purchases. This helps the business decide where to invest its budget next quarter, rather than treating every customer the same.


## Research Questions

I will be investigating the following questions:

1. The main question - **Does higher per-customer marketing spend lead to higher purchase amounts?**

2. More specifically, I aim to examine whether there is a meaningful positive relationship between *marketing_spend* and *purchase_amount* across the 150 customers in this campaign. 

3. I will also explore whether this relationship differs across customer segments (Consumer, SMB, and Enterprise), as different segment types might respond to marketing spend very differently.

## Research Question

The central question examined here is: **does higher per-customer marketing spend lead to higher purchase amounts?** More specifically, I want to know whether there is a meaningful positive relationship between `marketing_spend` and `purchase_amount` across the 150 customers in this campaign. I will also explore whether this relationship differs across customer segments (Consumer, SMB, and Enterprise), since different segment types might respond to marketing spend very differently.

## Dataset Introduction

The dataset records individual-level outcomes from a single quarter of a personalised marketing campaign. Each row represents one customer, capturing how much was spent reaching them and how much they ultimately purchased. The data was sourced from marketing campaign records and contains 150 customers with seven variables each.

```{r}
#| label: data-fetching

campaign_data_raw <- read_csv("data/36880817_Inayat Gill.csv", show_col_types = FALSE)
view(campaign_data_raw)

```

Before making any analysis, the data needs to be cleaned and be in a tidy format for easy and accurate interpretations. 

```{r}
#| label: clean-data

campaign_data_filter <- campaign_data_raw |>
  mutate(
    signup_date_clean = parse_date_time(
      signup_date,
      orders = c("Ymd", "bdy", "dmY"),
      quiet  = TRUE
    )
  ) |>
  filter(
    !is.na(signup_date_clean),  
    !is.na(purchase_amount)      
  ) |>
  select(
    customer_id,
    signup_date  = signup_date_clean,
    marketing_spend,
    visits,
    purchase_amount,
    segment
  )
```

The following changes were made to clean the data:

- The `signup_date` column contained dates in three inconsistent formats (ISO `YYYY-MM-DD`, written `Mon DD, YYYY`, and `DD/MM/YYYY`), which were unified into a single date format using `lubridate::parse_date_time()`. 

- The rows where `signup_date` could not be parsed were dropped. The rows having missing `purchase_amount` values (customers who did not make any purchase during the campaign period); were also filtered, as including them would distort the relationship between spend and purchase. 

- The `notes` column — which contained sparse labels like "contacted" or "returned" — was not used in the analysis, as it was too incomplete to be analytically useful. 

After cleaning, the dataset contained **123 complete observations** ready for analysis.

### Dataset Description
```{r}
#| label: dataset-size

cat("The filtered number of observations are :", nrow(campaign_data_filter), "\n") 
cat("The filtered number of variables are :", ncol(campaign_data_filter))

names(campaign_data_filter)
```

After cleaning, the dataset contains 124 observations 6 variables. 
The variables are: a customer id, signup date, marketing spend (in dollars), number of site visits, purchase amount (in dollars), and customer segment.

```{r}
#| label: tbl-summary
#| tbl-cap: Summary statistics for the three key numeric variables after cleaning.

summary_table <- data.frame(
  Statistic = c("Min", "Median", "Mean", "Max", "SD"),
  Marketing_Spend= c(
    min(campaign_data_filter$marketing_spend),
    median(campaign_data_filter$marketing_spend),
    mean(campaign_data_filter$marketing_spend),
    max(campaign_data_filter$marketing_spend),
    sd(campaign_data_filter$marketing_spend)
  ),
  Site_Visits = c(
    min(campaign_data_filter$visits),
    median(campaign_data_filter$visits),
    mean(campaign_data_filter$visits),
    max(campaign_data_filter$visits),
    sd(campaign_data_filter$visits)
  ),
  Purchase_Amount = c(
    min(campaign_data_filter$purchase_amount),
    median(campaign_data_filter$purchase_amount),
    mean(campaign_data_filter$purchase_amount),
    max(campaign_data_filter$purchase_amount),
    sd(campaign_data_filter$purchase_amount)
  ))

# Round all numbers to 2 decimal places
summary_table[, 2:4] <- round(summary_table[, 2:4], 2)

# Display as a table
kable(summary_table, align = "lrrr") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Marketing spend ranged from just $1.30 to $299.61, showing that customers received very different levels of investment during the campaign. Purchase amounts followed a similar spread, from $0 to $318.92, with an average of around $135. On average, customers visited the site about 6 times, suggesting reasonable engagement across the board.

```{r}
#| label: fig-segment-dist
#| fig-cap: "Distribution of marketing spend by customer segment"
#| fig-width: 6
#| fig-height: 4

ggplot(campaign_data_filter, aes(x = segment, y = marketing_spend, fill = segment)) +
  geom_boxplot() +
  labs(
    x = "Customer Segment",
    y = "Marketing Spend (USD)"
  ) +
     theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

@fig-segment-dist shows that marketing spend was distributed fairly broadly within each segment. Enterprise customers saw the widest variability, which makes sense given the fewer customers and more tailored outreach. Consumer and SMB segments received more uniform treatment. There is no obvious systematic difference in spend levels across the three groups, which is useful — it means spend variation is not simply a proxy for segment membership.

## Results

```{r}
#| label: compute-correlation

r_label <- round(cor(campaign_data_filter$marketing_spend, campaign_data_filter$purchase_amount), 2)

# Linear regression
fit    <- lm(purchase_amount ~ marketing_spend, data = campaign_data_filter)
slope  <- round(coef(fit)[2], 2)
r2     <- round(summary(fit)$r.squared, 3)
```

```{r}
#| label: fig-scatter
#| fig-cap: Relationship between per-customer marketing spend and purchase amount.
#| fig-width: 7
#| fig-height: 5

ggplot(campaign_data_filter, aes(x = marketing_spend, y = purchase_amount, colour = segment)) +
  geom_point() +
    geom_smooth(method = "lm", se = TRUE, colour = "grey30", linewidth = 0.9) +
  labs(
    x      = "Marketing Spend (USD)",
    y      = "Purchase Amount (USD)",
    colour = "Segment"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

```

```{r}
#| label: fig-facet
#| fig-cap: Scatter plots of marketing spend versus purchase amount, split by customer segment
#| fig-width: 8
#| fig-height: 4
ggplot(campaign_data_filter, aes(x = marketing_spend, y = purchase_amount, colour = segment)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ segment) +
  labs(
    x      = "Marketing Spend (USD)",
    y      = "Purchase Amount (USD)",
    colour = "Segment"
  ) +
     theme_minimal(base_size = 12) +
  theme(legend.position = "none")

```

